---
title: "Fall-of-Wicket Analysis"
author: "L. Blake"
date: "24/03/2021"
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/notebooks") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, tidymodels, vip, survival, ggfortify)

bbb <- readRDS("../../data/processed/bbb_cleaned.RDS") %>% 
  select(-c(bat_team_total_runs, 
            bat_team_total_wkts, bowl_team_total_runs, bowl_team_total_wkts,
            host_country, venue, winner, margin, outcome, toss_win, toss_elect,
            batter, bowler, dism_mode, bat_win_toss, 
            bat_home_away, bat_team, bowl_team, runs, extras, spell_balls,
            spell_runs, spell_wkts, pitch_factor, start_date)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate_at(c("innings", "bat_position"), as.factor)

# Normalise pitch factors
bbb <- bbb %>% mutate(rsum = seam_factor + spin_factor) %>% mutate(seam_factor = seam_factor/rsum) %>%
  mutate(spin_factor = spin_factor/rsum) %>% select(-rsum)

```




# Exploratory Data Analysis

```{r}
skimr::skim(bbb)
```

We shall first explore the patterns in the predictors when the wicket itself falls.

```{r}
wkt_only <- bbb %>% filter(is_wkt == "W")
```



```{r}
wkt_only %>% ggplot(aes(bat_balls)) + geom_density()

```
However, the pattern here is primarily due to the scarcity of the data. Let's try a relative wicket-likelihood.

```{r}
gb_wkt_freq <- function(df, col) {
  return(df %>% group_by_(col) %>%
  summarise(n_wkt = sum(is_wkt == "W"), n_no = sum(is_wkt == "no")) %>%
  mutate(wkt_freq = (n_wkt + 1) / (n_wkt + n_no + 1)) %>%
  mutate(logit = log(wkt_freq/(1 - wkt_freq))))    # With a pseudocount to prevent division by 0
}

  gb_wkt_freq(bbb, "bat_balls") %>%
  filter(bat_balls <= 300) %>%
  ggplot(aes(x = bat_balls, y = logit)) + 
  geom_point() +
  geom_smooth()
```

```{r}
gb_wkt_freq(bbb %>% mutate(bin = ntile(bat_avg, 20)), "bin") %>%
  ggplot(aes(x = bin, y = logit)) + 
  geom_point() +
  geom_smooth()
```


```{r}
gb_wkt_freq(bbb %>% mutate(bin = ntile(bowl_avg, 20)), "bin") %>%
  ggplot(aes(x = bin, y = logit)) + 
  geom_point() +
  geom_smooth()
```


```{r}
gb_wkt_freq(bbb %>% mutate(bin = ntile(team_lead, 20)), "bin") %>%
  ggplot(aes(x = bin, y = logit)) + 
  geom_point() +
  geom_smooth()
```

```{r}
gb_wkt_freq(bbb %>% mutate(bin = ntile(spin_factor, 20)), "bin") %>%
  ggplot(aes(x = bin, y = logit)) + 
  geom_point() +
  geom_smooth()
```

```{r}
gb_wkt_freq(bbb, "bat_score") %>%
  filter(bat_score < 250) %>%
  ggplot(aes(x = bat_score, y = logit)) + 
  geom_point() +
  geom_smooth()
```

```{r}
gb_wkt_freq(bbb, "inn_balls") %>%
  filter(inn_balls < 80*6) %>%
  ggplot(aes(x = inn_balls, y = wkt_freq)) + 
  geom_point() +
  geom_smooth()
```

```{r}
gb_wkt_freq(bbb, "inn_balls") %>%
  filter(80*6 <= inn_balls & inn_balls <= 160*6) %>%
  ggplot(aes(x = inn_balls, y = wkt_freq)) + 
  geom_point() +
  geom_smooth()
```


```{r}
gb_wkt_freq(bbb %>% mutate(bin = ntile(bowl_sr, 20)), "bin") %>%
  ggplot(aes(x = bin, y = wkt_freq)) + 
  geom_point() +
  geom_smooth()
```


## Bivariate Scatterplots

```{r}
wkt_only %>% ggplot(aes(x = bat_avg, y = bowl_avg)) + geom_point()

```



## By categorical variables

```{r}
wkt_only %>% ggplot(aes(x = bat_position))
```



# Training/Testing split
Split the data into training and testing sets, with a 75/25 split.
```{r}
set.seed(2403)
split <- initial_split(bbb, strata = is_wkt)
train <- training(split)
test <- testing(split)

```

Wickets are relatively rare in the full dataset, so let's create a balanced training set, with a 1:5 ratio.

```{r}
train_bal <- bind_rows(train %>% filter(is_wkt == "W"), sample_n(train %>% filter(is_wkt == "no"), size = 5*nrow(train %>% filter(is_wkt == "W"))))

```



# Model 2 - Gamma GLM
Driven by this result, some sort of exponential family distribution seems appropriate, when considering the number of balls faced before dismissal.
```{r}
wkt_only %>% ggplot(aes(bat_balls)) + geom_density()
```



# Model 1 - Logistic Regression


We first try fitting a logistic regression with every predictor in a linear relationship
```{r}
wf_log_basic <- workflow() %>%
  add_recipe(recipe(is_wkt ~ ., data = train_bal) %>%
               step_rm(bowl_class) %>%
               step_naomit(all_predictors()) %>%
               step_normalize(all_numeric()) %>%
               step_dummy(innings, bat_position, bat_arm, bowl_type)) %>%
  add_model(logistic_reg() %>% set_engine("glm"))
wf_log_basic <- wf_log_basic %>% fit(data = train_bal)
fit_log_basic <- wf_log_basic %>% pull_workflow_fit()
```

```{r}
tidy(fit_log_basic)

```

## Evaluate on Test Set

```{r}
prep_test_log_basic <- wf_log_basic %>% pull_workflow_prepped_recipe() %>% bake(new_data = test)

test_log_basic <- prep_test_log_basic %>% add_column(wf_log_basic %>% predict(new_data = test, type = "prob"))
test_log_basic %>% roc_curve(truth = is_wkt, estimate = .pred_W, event_level = "second") %>% autoplot()
test_log_basic %>% roc_auc(truth = is_wkt, estimte = .pred_W, event_level = "second") 


#dec_log <- train_bal %>%
#  add_column(log_fit %>% predict(new_data = train_bal %>% mutate(bat_score_int = bat_score), type = "prob")) %>%
#  mutate(logit = log(.pred_yes/(1-.pred_yes)))
```

## Variable impoirtance
```{r}
fit_log_basic %>% vi() %>%
  ggplot(aes(Importance, fct_reorder(Variable, Importance), fill = Sign)) + 
  geom_col()
```


Next, we try a model with less predictors and an expected interaction between the bowling type and the pitch conditions.
```{r}
wf_log_less <- workflow() %>%
  add_recipe(recipe(is_wkt ~ ., data = train_bal) %>%
               step_rm(bowl_wkts, team_score, bat_position, career_bowl_balls, bowl_type, seam_factor) %>%
               step_naomit(all_predictors()) %>%
               step_dummy(innings, bat_arm, bowl_class) %>%
               step_interact(terms = ~starts_with("bowl_class"):spin_factor)) %>%
  add_model(logistic_reg() %>% set_engine("glm"))
wf_log_less <- wf_log_less %>% fit(data = train_bal)
fit_log_less <- wf_log_less %>% pull_workflow_fit()
```

```{r}
tidy(fit_log_less)

```

## Evaluate on Test Set

```{r}
prep_test_log_less <- wf_log_less %>% pull_workflow_prepped_recipe() %>% bake(new_data = test)

test_log_less <- prep_test_log_less %>% add_column(wf_log_less %>% predict(new_data = test, type = "prob"))
test_log_less %>% roc_curve(truth = is_wkt, estimate = .pred_W, event_level = "second") %>% autoplot()
test_log_less %>% roc_auc(truth = is_wkt, estimte = .pred_W, event_level = "second") 
```


## Variable impoirtance
```{r}
fit_log_less %>% vi() %>%
  ggplot(aes(Importance, fct_reorder(Variable, Importance), fill = Sign)) + 
  geom_col()
```


## Assumption checkiong
```{r}
train_log_basic <- train_bal %>% na.omit() %>%
  add_column(wf_log_basic %>% predict(new_data = train_bal, type = "prob")) %>%
  mutate(logit = log(.pred_W/(1-.pred_W)))
```

### Linearity
```{r}
# Linearity
ggplot(aes(bat_avg, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(bowl_avg, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(bat_score, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(team_score, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(bat_balls, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(bowl_balls, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(team_lead, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(match_balls, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

```



# Random Forest

```{r}
rf_wf <- workflow() %>%
  add_recipe(recipe(is_wkt ~ ., data = train_bal) %>% 
               step_naomit(everything()) %>%
               step_rm(c(bowl_class, game_id, career_bat_balls, career_bowl_balls))) %>%
  add_model(rand_forest(mode="classification", trees=300) %>% 
  set_engine("ranger", importance="permutation"))
  
rf_wf <- rf_wf %>% fit(data = train_bal)

(rf_fit <- rf_wf %>% pull_workflow_fit())
```

## Evaluate on test set
```{r}
test_rf_prep <- rf_wf %>% pull_workflow_prepped_recipe() %>% bake(new_data = test)

test_rf <- test %>% na.omit() %>% add_column(rf_fit %>% predict(new_data = test_rf_prep, type = "prob"))
test_rf %>% roc_curve(truth = is_wkt, estimate = .pred_W, event_level = "second") %>% autoplot()
test_rf %>% roc_auc(truth = is_wkt, estimte = .pred_W, event_level = "second") 
```

## Importance
```{r}
rf_fit %>% vi() %>% 
  ggplot(aes(Importance, fct_reorder(Variable, Importance))) + 
  geom_col() + 
  labs(y = NULL)
```


# Survival Analysis
Let us instead put this in the context of survival analysis: we want to predict the probability that a batter survives beyond a certain number of deliveries.


```{r}
bbb %>% group_by(bat_balls) %>% summarize(p = sum(is_wkt == "W")/n()) %>%
  filter(p != 0) %>% filter(bat_balls < 300) %>%
  ggplot(aes(x = bat_balls, y = p)) + geom_point()
```

## Exponential GLM
```{r}
form_exp <- bat_balls ~ bat_avg + bowl_avg + bat_sr
recp_exp <- recipe(form_exp, data = train %>% filter(is_wkt == "W")) %>%
  step_naomit(everything()) %>%
  step_normalize(all_numeric()) %>%
  step_mutate(bat_balls = bat_balls + 1)%>%
  #step_mutate(is_wkt = as.numeric(is_wkt) - 1) %>%
  prep(data = train)


fit_exp <- glm(form_exp, family = Gamma(link = "log"), data = recp_exp %>% juice())
summary(fit_exp, dispersion = 1)
```


### Evaluate on testing set

```{r}
prep_test_exp <- recp_exp %>% bake(new_data = test %>% filter(is_wkt == "W"))

test_exp <- prep_test_exp %>% add_column(rate = fit_exp %>% predict(newdata = prep_test_exp, type = "terms", dispersion = 1))
test_exp %>% roc_curve(truth = is_wkt, estimate = .pred_W, event_level = "second") %>% autoplot()
test_exp %>% roc_auc(truth = is_wkt, estimte = .pred_W, event_level = "second") 
```



## Kaplan-Meier Analysis
```{r}
bbb_01 <- bbb %>% mutate(is_wkt = as.numeric(is_wkt) - 1)

km_fit <- survfit(Surv(bat_balls, is_wkt) ~ 1, data = bbb_01)
autoplot(km_fit)
```


### By factors
```{r}
km_inn_fit <- survfit(Surv(bat_balls, is_wkt) ~ innings, data = bbb_01)
autoplot(km_inn_fit)
```


## Cox Proportional Model
```{r}
train_01 <- train %>% mutate(is_wkt = as.numeric(is_wkt) - 1)
fit_cox <- coxph(Surv(bat_balls, is_wkt) ~ bat_avg + bowl_avg + bowl_sr + bat_score + team_lead + match_balls, data = train_01)
summary(fit_cox)
autoplot(survfit(fit_cox))
```

### Proportional hazards assumption
```{r}
zp <- cox.zph(fit_cox)
```

```{r}
plot(zp[1])
```


