---
title: "Fall-of-Wicket Analysis"
author: "L. Blake"
date: "24/03/2021"
output: html_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../output/notebooks") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, tidymodels, vip)

bbb <- readRDS("../../data/processed/bbb_cleaned.RDS") %>% 
  select(-c(bat_team_total_runs, 
            bat_team_total_wkts, bowl_team_total_runs, bowl_team_total_wkts,
            host_country, venue, winner, margin, outcome, toss_win, toss_elect,
            batter, bowler, dism_mode, bat_win_toss, 
            bat_home_away, bat_team, bowl_team, runs, extras, spell_balls,
            spell_runs, spell_wkts, pitch_factor, start_date)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate_at(c("innings", "bat_position"), as.factor)

# Normalise pitch factors
bbb <- bbb %>% mutate(rsum = seam_factor + spin_factor) %>% mutate(seam_factor = seam_factor/rsum) %>%
  mutate(spin_factor = spin_factor/rsum) %>% select(-rsum)

```



# Exploratory Data Analysis

```{r}
skimr::skim(bbb)
```

We shall first explore the patterns in the predictors when the wicket itself falls.

```{r}
wkt_only <- bbb %>% filter(is_wkt == "W")
```



## Bivariate Scatterplots

```{r}
wkt_only %>% ggplot(aes(x = bat_avg, y = bowl_avg)) + geom_point()

```


# Training/Testing split
Split the data into training and testing sets, with a 75/25 split.
```{r}
set.seed(2403)
split <- initial_split(bbb, strata = is_wkt)
train <- training(split)
test <- testing(split)

```

Wickets are relatively rare in the full dataset, so let's create a balanced training set, with a 1:5 ratio.

```{r}
train_bal <- bind_rows(train %>% filter(is_wkt == "W"), sample_n(train %>% filter(is_wkt == "no"), size = 5*nrow(train %>% filter(is_wkt == "W"))))

```



# Model 1 - Logistic Regression


We first try fitting a logistic regression with every predictor in a linear relationship
```{r}
wf_log_basic <- workflow() %>%
  add_recipe(recipe(is_wkt ~ ., data = train_bal) %>%
               step_rm(bowl_class) %>%
               step_naomit(all_predictors()) %>%
               step_normalize(all_numeric()) %>%
               step_dummy(innings, bat_position, bat_arm, bowl_type)) %>%
  add_model(logistic_reg() %>% set_engine("glm"))
wf_log_basic <- wf_log_basic %>% fit(data = train_bal)
fit_log_basic <- wf_log_basic %>% pull_workflow_fit()
```

```{r}
tidy(fit_log_basic)

```

## Evaluate on Test Set

```{r}
prep_test_log_basic <- wf_log_basic %>% pull_workflow_prepped_recipe() %>% bake(new_data = test)

test_log_basic <- prep_test_log_basic %>% add_column(wf_log_basic %>% predict(new_data = test, type = "prob"))
test_log_basic %>% roc_curve(truth = is_wkt, estimate = .pred_W, event_level = "second") %>% autoplot()
test_log_basic %>% roc_auc(truth = is_wkt, estimte = .pred_W, event_level = "second") 


#dec_log <- train_bal %>%
#  add_column(log_fit %>% predict(new_data = train_bal %>% mutate(bat_score_int = bat_score), type = "prob")) %>%
#  mutate(logit = log(.pred_yes/(1-.pred_yes)))
```

## Variable impoirtance
```{r}
fit_log_basic %>% vi() %>%
  ggplot(aes(Importance, fct_reorder(Variable, Importance), fill = Sign)) + 
  geom_col()
```


Next, we try a model with less predictors and an expected interaction between the bowling type and the pitch conditions.
```{r}
wf_log_less <- workflow() %>%
  add_recipe(recipe(is_wkt ~ ., data = train_bal) %>%
               step_rm(bowl_wkts, team_score, bat_position, career_bowl_balls, bowl_type, seam_factor) %>%
               step_naomit(all_predictors()) %>%
               step_dummy(innings, bat_arm, bowl_class) %>%
               step_interact(terms = ~starts_with("bowl_class"):spin_factor)) %>%
  add_model(logistic_reg() %>% set_engine("glm"))
wf_log_less <- wf_log_less %>% fit(data = train_bal)
fit_log_less <- wf_log_less %>% pull_workflow_fit()
```

```{r}
tidy(fit_log_less)

```

## Evaluate on Test Set

```{r}
prep_test_log_less <- wf_log_less %>% pull_workflow_prepped_recipe() %>% bake(new_data = test)

test_log_less <- prep_test_log_less %>% add_column(wf_log_less %>% predict(new_data = test, type = "prob"))
test_log_less %>% roc_curve(truth = is_wkt, estimate = .pred_W, event_level = "second") %>% autoplot()
test_log_less %>% roc_auc(truth = is_wkt, estimte = .pred_W, event_level = "second") 
```


## Variable impoirtance
```{r}
fit_log_less %>% vi() %>%
  ggplot(aes(Importance, fct_reorder(Variable, Importance), fill = Sign)) + 
  geom_col()
```


## Assumption checkiong
```{r}
train_log_basic <- train_bal %>% na.omit() %>%
  add_column(wf_log_basic %>% predict(new_data = train_bal, type = "prob")) %>%
  mutate(logit = log(.pred_W/(1-.pred_W)))
```

### Linearity
```{r}
# Linearity
ggplot(aes(bat_avg, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(bowl_avg, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(bat_score, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(team_score, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(bat_balls, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(bowl_balls, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(team_lead, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

ggplot(aes(match_balls, logit), data = train_log_basic) + geom_point()  +
  geom_smooth(method = "loess")

```


