---
title: "The Naive FOW Model"
author: "L. Blake"
date: "26/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, tidymodels)
source("../pitch_qi.R")

bbb <- readRDS("../../data/processed/bbb_cleaned.RDS") %>%
  mutate_if(is.character, as.factor) %>%
  mutate_at(c("innings", "bat_position"), as.factor)

# Use better pitch factors:
bbb <- pqi_on_all(bbb) %>%
  # Wicket cases should already be handled by a prior model
  select(-c(
    bat_team_total_runs,
    bat_team_total_wkts, bowl_team_total_runs, bowl_team_total_wkts,
    host_country, venue, winner, margin, outcome, toss_win, toss_elect,
    batter, bowler, dism_mode, bat_win_toss,
    bat_home_away, bat_team, bowl_team, spell_balls, extras,
    spell_runs, spell_wkts, pitch_factor, start_date, game_id, bowl_class, bowl_wkts, bowl_runs, spin_factor, seam_factor
  )) #%>%
  #na.omit()

# Combine left and right-arm medium into one level
bbb <- bbb %>% mutate(
  bowl_type = fct_recode(bowl_type, m = "rm", m = "lm")
)
```

We attempt to predict the probability of a wicket falling, by inspecting the relationship between each covariate and the log-odds of a wicket falling and choosing an appropriate transformation. These are used in a logistic model to provide a simple benchmark which is easy to implement. 


# Summary Statistics

```{r}
skimr::skim(bbb)
```

# Calculating the log-odds

```{r}
gb_wkt_freq <- function(df, col) {
  return(df %>% group_by_(.dots = col) %>%
    summarise(n_wkt = sum(is_wkt == "W"), n_no = sum(is_wkt == "no")) %>%
    mutate(wkt_freq = (n_wkt + 1) / (n_wkt + n_no + 1)) %>% # With a pseudocount to prevent division by 0
    mutate(logit = log(wkt_freq / (1 - wkt_freq))))
}
```


# Each possible predictor
## The number of deliveries
```{r}
gb_wkt_freq(bbb, "bat_balls") %>%
  filter(bat_balls <= 300) %>%
  ggplot(aes(x = bat_balls, y = logit)) +
  geom_point() +
  geom_smooth()
```

## The number of balls in innings
```{r}
gb_wkt_freq(bbb, "inn_balls") %>%
  filter(inn_balls < 600) %>%
  ggplot(aes(x = inn_balls, y = logit)) +
  geom_point() +
  geom_smooth()
```


When the new ball is taken after 80 overs, this is a discrete event 
```{r}
bbb <- bbb %>%
  mutate(ball = ifelse(inn_balls > 0, inn_balls %/% 480, 0), ball_age = inn_balls %% 480) %>%
  mutate_at(c("ball"), as.factor) %>%
  select(-inn_balls)
```

```{r}
gb_wkt_freq(bbb, "ball_age") %>%
  ggplot(aes(x = ball_age, y = logit)) +
  geom_point() +
  geom_smooth()
```

```{r}
gb_wkt_freq(bbb, "ball") %>%
  ggplot(aes(x = ball, y = logit)) +
  geom_point() +
  geom_smooth()
```

```{r}
gb_wkt_freq(bbb, c("ball", "ball_age")) %>%
  ggplot(aes(x = ball_age, y = logit, color = ball)) +
  geom_point() 
```


## Team Wickets
```{r}
gb_wkt_freq(bbb, "team_wkts") %>%
  ggplot(aes(x = team_wkts, y = logit)) +
  geom_point() +
  geom_smooth()
```
However, this is a misleading picture.

```{r}
bbb %>%
  filter(career_bat_balls > 1000) %>%
  ggplot(aes(x = team_wkts, y = bat_avg)) +
  geom_boxplot()
```


## Team lead
```{r}
gb_wkt_freq(bbb, "team_lead") %>%
  ggplot(aes(x = team_lead, y = logit)) +
  geom_point() +
  geom_smooth()
```


## Batting average
```{r}
gb_wkt_freq(bbb %>% mutate(bin = ntile(bat_avg, 100)), "bin") %>%
  ggplot(aes(x = bin, y = logit)) +
  geom_point() +
  geom_smooth()
```


## Batting Dismissal Frequency
```{r}
bbb <- bbb %>%
  mutate(bat_dism_freq = bat_avg / (100 * bat_sr)) %>%
  na.omit() # TODO: handle zero strike rate
gb_wkt_freq(bbb %>% mutate(bin = ntile(bat_dism_freq, 100)), "bin") %>%
  ggplot(aes(x = bin, y = logit)) +
  geom_point() +
  geom_smooth()
```

## Bowling strike rate
```{r}
gb_wkt_freq(bbb %>% mutate(bin = ntile(bowl_sr, 100)), "bin") %>%
  ggplot(aes(x = bin, y = logit)) +
  geom_point() +
  geom_smooth()
```


# Training/Testing split
```{r}
set.seed(2403)
split <- initial_split(bbb, strata = is_wkt)
train <- training(split)
test <- testing(split)
```


# The model

```{r}
rec <- recipe(is_wkt ~ ., data = train) %>%
  step_rm(team_score, bat_position, bat_avg, bowl_balls, bowl_avg, career_bat_balls, career_bowl_balls) %>%
  # These predictors are closely related to others
  step_poly(team_lead, degree = 3) %>%
  step_BoxCox(all_numeric()) %>% # To account for any skew
  step_dummy(team_wkts, bat_arm, bowl_type, ball) %>%
  step_interact(terms=~starts_with("bat_arm"):starts_with("bowl_type"))

wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(logistic_reg(penal
                         \
                         ty=0.5) %>% set_engine("glm"))

wf <- wf %>% fit(data = train)
fitted <- wf %>% pull_workflow_fit()
summary(fitted$fit)
```


94543

## Summary of Recipe
```{r}
prep_rec <- wf %>% pull_workflow_prepped_recipe()
for (n in 1:length(prep_rec$steps)) {
  print(tidy(prep_rec, n = n))
}
```
